---
title: "Final Project of MSDS 597"
author: "Hui Wang"
date: "5/4/2020"
output:
  pdf_document: default
  word_document: default
  html_document:
    df_print: paged
geometry: margin=1in
fontsize: 11pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height = 6)
options(tinytex.verbose = TRUE)
library(dplyr)
library(tidyverse)
library(tidytext)
library(lubridate)
library(xml2)
library(rvest)
library(stringr)
library(rvest)
library(gutenbergr)   # for the text 
library(tidyr)   # for separate_rows
library(gridExtra)    #to make side-by-side figure
library(choroplethr)
library(choroplethrMaps)
library(sf)
library(scales)   # for percent()
library(ggplot2)
```

(With Presentaion) Link in Github: https://github.com/Hui-Dora-Wang/Data-Wrangling-FinalProject

## COVID-19, Political Standing, Education Level, Population Density, and GDP

During this hard time, COVID-19 is a terrifying yet unavoidable keyword in our life. In this project, when applying the skills learned from the course, I try to find if there is any direct relationship between the spread of COVID-19 and the political standing, education level, population density, and GDP of the states in the US.

```{r COVID-19, warning=FALSE,echo = FALSE,fig.height=4,fig.width=6,fig.align="center"}
# import the data
COVID_19_raw <- "https://www.worldometers.info/coronavirus/country/us/" %>%
  read_html() %>%
  html_table(fill=TRUE) %>%
  .[[1]]
# head(COVID_19_raw)

# colnames(COVID_19)

# select the rows and columns needed
COVID_19 <- data.frame(COVID_19_raw)
COVID_19 <- COVID_19[-c(1,53:nrow(COVID_19)),]
COVID_19 <- COVID_19 %>%
  dplyr::select(USAState, TotalCases, TotalDeaths,Tot.Cases.1M.pop,Deaths.1M.pop,Tests.1M.pop) %>%
  arrange(USAState)

# rename the columns
names(COVID_19) <-  c("state","TotalCases","TotalDeaths","Cases.1M.pop","Deaths.1M.pop","Tests.1M.pop")

# change state name to lower letter
COVID_19$state <- tolower(COVID_19$state)

# remove comma and numerical
COVID_19$TotalCases <- as.numeric(gsub(",", "", COVID_19$TotalCases))
COVID_19$TotalDeaths <- as.numeric(gsub(",", "", COVID_19$TotalDeaths))
COVID_19$Cases.1M.pop <- as.numeric(gsub(",", "", COVID_19$Cases.1M.pop))
COVID_19$Deaths.1M.pop <- as.numeric(gsub(",", "", COVID_19$Deaths.1M.pop))
COVID_19$Tests.1M.pop <- as.numeric(gsub(",", "", COVID_19$Tests.1M.pop))

# fill NA with 0
COVID_19$TotalDeaths <- replace_na(COVID_19$TotalDeaths, 0)

# head(COVID_19)
```

To make the full dataset, I scraped five tables from different websites and did the data cleaning for four of them. Let’s use the Political Standing table as a simple example; it provides the data of the 2016 presidential election. 
As you can see, to process the raw data scraped from the website, we need to split the strings and convert the data type to numerical. And since the names of the states are abbreviations, we need to assign the names of states by hand. It is easier after arranging them by state and we just need to pay attention to some different orders between the states, such as the District of Columbia is after Delaware, but D.C. was in front of Del in the original table.

```{r Political Standing, warning=FALSE,echo = FALSE,fig.height=4,fig.width=6,fig.align="center"}
# import the data
political_raw <- "https://www.nytimes.com/elections/2016/results/president" %>%
  read_html() %>%
  html_table(fill=TRUE) %>%
  .[[2]]
head(political_raw)

# select the needed columns
# add the region names by hand, pay attendsion on such as district of colunmbia which in short is D.C. and stand after delaware, stand in front of it.
political <- political_raw %>%
  dplyr::select(State,Clinton,Trump,Other) %>%
  arrange(State) %>%
  unique() %>% 
  mutate(state = c("alabama","alaska","arizona","arkansas","california","colorado","connecticut","district of columbia","delaware","florida","georgia","hawaii","idaho","illinois","indiana","iowa","kansas","kentucky","louisiana","massachusetts","maryland","maine","michigan","minnesota","mississippi","missouri","montana","north carolina","north dakota","new hampshire","new jersey","new mexico","new york","nebraska","nevada","ohio","oklahoma","oregon","pennsylvania","rhode island","south carolina","south dakota","tennessee","texas","utah","virginia","vermont","west virginia","washington","wisconsin","wyoming")) %>%
  dplyr::select(state,Clinton,Trump,Other)

# extracting the digit numbers and then divided by 100
political$Clinton <- as.numeric(gsub("[^0-9]","",political$Clinton))/100
political$Trump <- as.numeric(gsub("[^0-9]","",political$Trump))/100
political$Other <- as.numeric(gsub("[^0-9]","",political$Other))/100

# change the columns' names
colnames(political) <- c("state","Clinton","Trump","Other_poli")

# tried the following, but it's not numueric anymore
# percent(as.numeric(gsub("[^0-9]","",Clinton))/100)

# sum = Clinton+Trump+Other_poli is not 1 actually, checked true

head(political)
```

```{r Education, warning=FALSE,echo = FALSE,fig.height=4,fig.width=6,fig.align="center"}
# import the data
education_raw <- "https://worldpopulationreview.com/states/most-educated-states/" %>%
  read_html() %>%
  html_table(fill=TRUE) %>%
  .[[1]]
# head(education_raw)   # already clean enough

# select the needed col
education <- education_raw %>%
  dplyr::select(State,`Education Score`) %>%
  arrange(State)

# change the clo names
names(education) <- c("state","Education")

# lower the state col
education$state <- tolower(education$state)

# head(education)
```

```{r Population Density, warning=FALSE,echo = FALSE,fig.height=4,fig.width=6,fig.align="center"}
# import the data
pop_density_raw <- "https://simple.wikipedia.org/wiki/List_of_U.S._states_by_population_density" %>%
  read_html() %>%
  html_table(fill=TRUE) %>%
  .[[4]]
# head(pop_density_raw)

pop_density <- pop_density_raw %>%
  dplyr::select(`State/Territory/Division/Region`,`2013 estimatedpopulation`,`Land area(sqmi)`,`Population/sqmi`) %>%
  arrange(`State/Territory/Division/Region`)

# change column name 
names(pop_density) <- c("state","Population","Land","Density") 

# delete the comma and numerical the colums
pop_density$Population <- as.numeric(gsub(",", "", pop_density$Population))
pop_density$Land <- as.numeric(gsub(",", "", pop_density$Land))
pop_density$Density <- as.numeric(gsub(",", "", pop_density$Density))

# change state name to lower letter
pop_density$state <- tolower(pop_density$state)

# check density
pop_density_check <- pop_density %>%
  mutate(den = pop_density$Population/pop_density$Land)

# head(pop_density)
```

```{r GDP, warning=FALSE,echo = FALSE,fig.height=4,fig.width=6,fig.align="center"}
# import the data
GDP_raw <- "https://en.wikipedia.org/wiki/List_of_U.S._states_and_territories_by_GDP" %>%
  read_html() %>%
  html_table(fill=TRUE) %>%
  .[[3]]
# head(GDP_raw)

# change cols name 
names(GDP_raw) <- c("rank","state","Q4_2019","GDP_portion","GDP_perCapita","Area")

# select the needed cols
GDP <- GDP_raw %>%
  dplyr::select(state,Q4_2019,GDP_portion,GDP_perCapita,Area) %>%
  arrange(state)

# delete the comma and numerical the colums
GDP$Q4_2019 <- as.numeric(gsub(",", "", GDP$Q4_2019))
GDP$GDP_portion <- as.numeric(GDP$GDP_portion)
GDP$GDP_perCapita <- as.numeric(gsub(",", "", GDP$GDP_perCapita))

# change state name to lower letter
GDP$state <- tolower(GDP$state)

# head(GDP)
```

## Full Table

Well, after cleaning up all the tables and left join them by state, we can get the full table. It contains 50 states and DC in row and 18 variables. The names of columns are pretty straightforward. 

```{r Full Table, warning=FALSE,echo = FALSE,fig.height=4,fig.width=6,fig.align="center"}
# join tables COVID_19 and political
full1 <- left_join(COVID_19,political,by="state")
# join table education and filled NA (district of columbia) with 0
full2 <- left_join(full1,education,by="state")
full2$Education <- replace_na(full2$Education, 0)
# join table pop_density
full3 <- left_join(full2,pop_density,by="state")
# join table pop_density
full <- left_join(full3,GDP,by="state")
head(full)

# export the .csv file
write_csv(full, "COVID19-Pol-Edu-Pop-GDP.csv")
```

## COVID-19, Political Standing, and GDP

Well, let’s see the plots showing the relationship of the spread of COVID-19, the Political Standing, and GDP of the states in America. As mentioned before, the Political Standing stands for whether more popular votes went for Clinton or Trump in 2016. The GDP is the GDP per capita of the fourth season in 2019. 
In the plots, we have logged GDP per capita on the x-axis and logged COVID-19 case number, death number and test number on the y-axis. The two lines represent either the blue states, that voted for Clinton, and the red states, that voted for Trump. 
In the case plot, we can see that both lines go upward because both GDP and COVID-19 spread are positively correlated with urbanization: urban areas with high population density. The difference between the two slopes may be due to the different test rates. 
From the test plot, although the difference between slopes is subtle, the blue states have conducted more tests than the red states at any given GDP level. 
The death plot is interesting. We would expect places with higher GDP to be more prepared for COVID-19 and thus have a lower death rate. Indeed, the line for Trump states is what we expect. The one for Clinton does not appear so probably because there are too many confirmed cases in those big cities and the medical system breaks down.

```{r COVID-19, Political Standing, and GDP, warning=FALSE,echo = FALSE,fig.height=4,fig.width=6,fig.align="center"}
# select the needed columns
# combine two Political Standing colums and define
full_ForMap1 <- full %>%
  dplyr::select(state,Cases.1M.pop,Deaths.1M.pop,Tests.1M.pop,Clinton,Trump,GDP_perCapita) %>%
  mutate(Poli_standing = case_when(
    Clinton > Trump ~ "Clinton",
    Clinton < Trump ~ "Trump"))

# four states are very close
# Trump got Michigan, Pennsylvania, Wisconsin; Clinton got New Hampshire
full_ForMap1$Poli_standing <- replace_na(full_ForMap1$Poli_standing,"Trump")
full_ForMap1$Poli_standing[30] <- "Clinton"

# log transformation
full_ForMap1$`Deaths.1M.pop(log)` <- log(full_ForMap1$Deaths.1M.pop)
full_ForMap1$`Cases.1M.pop(log)` <- log(full_ForMap1$Cases.1M.pop)
full_ForMap1$`GDP_perCapita(log)` <- log(full_ForMap1$GDP_perCapita)
full_ForMap1$`Tests.1M.pop(log)` <- log(full_ForMap1$Tests.1M.pop)

Clinton_Map <- full_ForMap1 %>%
  dplyr::select(state,`Cases.1M.pop(log)`,`Deaths.1M.pop(log)`,`Tests.1M.pop(log)`,Poli_standing,`GDP_perCapita(log)`) %>%
  filter(Poli_standing=="Clinton")

Trump_Map <- full_ForMap1 %>%
  dplyr::select(state,`Cases.1M.pop(log)`,`Deaths.1M.pop(log)`,`Tests.1M.pop(log)`,Poli_standing,`GDP_perCapita(log)`) %>%
  filter(Poli_standing=="Trump")

reg1 <- lm(`Deaths.1M.pop(log)` ~ `GDP_perCapita(log)`, data = Clinton_Map)
# summary(reg1)
reg2 <- lm(`Deaths.1M.pop(log)` ~ `GDP_perCapita(log)`, data = Trump_Map)
# summary(reg2)
reg3 <- lm(`Cases.1M.pop(log)` ~ `GDP_perCapita(log)`, data = Clinton_Map)
# summary(reg3)
reg4 <- lm(`Cases.1M.pop(log)` ~ `GDP_perCapita(log)`, data = Trump_Map)
# summary(reg4)
reg5 <- lm(`Tests.1M.pop(log)` ~ `GDP_perCapita(log)`, data = Clinton_Map)
# summary(reg5)
reg6 <- lm(`Tests.1M.pop(log)` ~ `GDP_perCapita(log)`, data = Trump_Map)
# summary(reg6)

par(mfrow = c(2,2))
ggplot()+
  geom_point(data=Clinton_Map, aes(x=`GDP_perCapita(log)`, y=`Cases.1M.pop(log)`), color = "blue", size=2) +
  geom_point(data=Trump_Map, aes(x=`GDP_perCapita(log)`, y=`Cases.1M.pop(log)`), color = "red", size=2) +
  stat_function(fun = function(`GDP_perCapita(log)`) -10.1383 + 1.6179*`GDP_perCapita(log)`, data=Clinton_Map, color = "blue") +
  stat_function(fun = function(`GDP_perCapita(log)`) 1.886 + 0.502*`GDP_perCapita(log)`, data=Trump_Map, color = "red") +
  ggtitle("COVID-19 Case, Political Standing, and GDP") +
  annotate("text", x = 12.2, y = 9.75, label = "Clinton", size=8, colour="blue") +
  annotate("text", x = 12.2, y = 8.2, label = "Trump", size=8, colour="red")

ggplot()+
  geom_point(data=Clinton_Map, aes(x=`GDP_perCapita(log)`, y=`Tests.1M.pop(log)`), color = "blue", size=2) +
  geom_point(data=Trump_Map, aes(x=`GDP_perCapita(log)`, y=`Tests.1M.pop(log)`), color = "red", size=2) +
  stat_function(fun = function(`GDP_perCapita(log)`) 5.6436 + 0.4006*`GDP_perCapita(log)`, data=Clinton_Map, color = "blue") +
  stat_function(fun = function(`GDP_perCapita(log)`) 5.6635 + 0.3823*`GDP_perCapita(log)`, data=Trump_Map, color = "red") +
  ggtitle("COVID-19 Test, Political Standing, and GDP") +
  annotate("text", x = 12.2, y = 10.6, label = "Clinton", size=8, colour="blue") +
  annotate("text", x = 12.2, y = 10.2, label = "Trump", size=8, colour="red")

ggplot() +
  geom_point(data=Clinton_Map, aes(x=`GDP_perCapita(log)`, y=`Deaths.1M.pop(log)`), color = "blue", size=2) +
  geom_point(data=Trump_Map, aes(x=`GDP_perCapita(log)`, y=`Deaths.1M.pop(log)`), color = "red", size=2) +
  geom_abline(intercept = -16.068, slope = 1.874, color = "blue") +
  geom_abline(intercept = 13.003, slope = -0.829, color = "red") +
  ggtitle("COVID-19 Death, Political Standing, and GDP") +
  annotate("text", x = 12.2, y = 7, label = "Clinton", size=8, colour="blue") +
  annotate("text", x = 12.2, y = 3, label = "Trump", size=8, colour="red")
```

## Education & Political Standing

I also want to see the relationship between the education level and political standing. And it seems the states with higher education level are more likely to have voted for Clinton instead of Trump. 

```{r Education & Political Standing, warning=FALSE,echo = FALSE,fig.height=4,fig.width=6,fig.align="center"}
# education for map
education_ForMap <- full %>%
  dplyr::select(state, Education)
# match with choroplethr package
# change columns' names
names(education_ForMap) <- c("region", "value")

# political standing in Clinton for map
political_Clinton_ForMap <- full %>%
  dplyr::select(state, Clinton)
## match with choroplethr package
# change column name 
names(political_Clinton_ForMap) <- c("region", "value")

# political standing in Trump for map
political_Trump_ForMap <- full %>%
  dplyr::select(state, Trump)
## match with choroplethr package
# change column name 
names(political_Trump_ForMap) <- c("region", "value")

par(mfrow = c(2,2))
# create three side-by-side choropleth maps of Education & Political Standing in the USA
data(continental_us_states)
grid.arrange(state_choropleth(education_ForMap, title  = "Education Level in the US", zoom = continental_us_states))
grid.arrange(state_choropleth(political_Clinton_ForMap, title  = "Vote for Clinton in 2016 in the US", zoom = continental_us_states),
             state_choropleth(political_Trump_ForMap, title  = "Vote for Trump in 2016 in the US", zoom = continental_us_states))
```

## COVID-19, Population Density, and GDP

Last, about the COVID-19, Population Density, and GDP, I delete the outlier, District of Columbia. Seeing the pairs plot, I think it would be interesting to compare the relationship between cases on one hand and population density or the GDP per capita on the other. And from the linear plots, we can see the model of population density in the middle plot fits better than the one of GDP per capita on the right plot. So, the GDP per capita is not that much related to the cases as population density is.

```{r COVID-19, Population Density, and GDP, warning=FALSE,echo = FALSE,fig.height=4,fig.width=6,fig.align="center"}
# delete the outlier 
density_ForMap <- full %>%
  filter(state != "district of columbia") %>%
  dplyr::select(Density,GDP_perCapita,Deaths.1M.pop,Cases.1M.pop)

par(mfrow = c(2,2))
pairs(density_ForMap)
ggplot(density_ForMap, aes(Density, Cases.1M.pop)) +
  geom_point() +
  geom_smooth(method='lm', formula = y~x) +
  ggtitle("COVID-19 Cases & Population Density")

ggplot(density_ForMap, aes(GDP_perCapita, Cases.1M.pop)) +
  geom_point() +
  geom_smooth(method='lm', formula = y~x) +
  ggtitle("COVID-19 Cases & GDP_perCapita")
```
